Value-Suppressing Uncertainty Maps

### 2017-12-08 paper/figures.tex
Larger figure
Added poll example

\newcommand{\taskTwoQuartiles}{
	\begin{figure}
		\centering
		\includegraphics[width=0.75\columnwidth]{exp2histogram.pdf}
		\caption{2D histogram of number of guesses from the prediction experiment for the cases when participants saw a traditional bivariate maps (left) and VSUPs (right). For each trial, participants made 5 guesses. While participants generally favored low uncertainty in guesses, our stimuli design encouraged them to make at least some guesses with higher uncertainty. VSUPs encouraged more guesses with middling uncertainty, and discouraged highly uncertain guesses.}
		\label{fig:taskTwoQuartiles}
	\end{figure}
}
%\newcommand{\performanceFig}{
%	\begin{figure}[t]
%		\centering
%		\includegraphics[width=0.9\columnwidth]{performance.pdf}
%		\caption{
%			The number of discrete color categories in traditional bivariate color maps versus VSUPs, assuming value is encoded by the Viridis color map~\protect\cite{viridis}, and uncertainty is encoded by saturation/value (whiter values are more uncertain). Since these whiter uncertain colors are closer together in CIELAB space, and VSUPs intentionally require fewer color bins in those regions, VSUPs always have more color categories than the corresponding 2D bivariate map. The red lines denote:
%		  \textbf{1)} Theoretical 50\% JND from CIELAB specification.
%			\textbf{2)} 50\% JND from controlled lab study of Mahy et al.~\protect\cite{mahy1994evaluation}.
%			\textbf{3)} 50\% JND from Szafir et al.\protect\cite{szafir2014adapting}.
%			\textbf{4)} 50\% JND for small visual targets from Stone et al.~\protect\cite{stone2014engineering}.
%			\textbf{5)} The threshold resulting in the simplest possible (2x2) bivariate map.
%	    }
%		\label{fig:performance}
%	\end{figure}
%}

@@

\newcommand{\pollFig}{
	\begin{figure}[t]
		\centering
		\includegraphics[width=\columnwidth]{polling-2d.pdf}
		\vspace{-15px}
		\caption{Polling data prior to the 2016 U.S. Presidential Election, encoded in a traditional 2D bivariate map. Many states appear to be ``leaning'' towards Clinton, even when the margin of error of the polls is quite large. The blue-red color scale that is semantically appropriate for the topic is also not isoluminant, making it difficult to see just how much uncertainty there is in states with small leads for either candidate.}
		\label{fig:polling2d}

@@

		\centering
		\includegraphics[width=\columnwidth]{polling-vsum.pdf}
		\vspace{-15px}
		\caption{Polling data prior to the 2016 U.S. Presidential Election, encoded with a VSUP. The larger amount of color bins for high-certainty regions shows that the leads for Hillary Clinton were, in many states, quite small. The collapsing of bins as uncertainty increases dissuades viewers from making confident predictions about states where the outcome was very much in doubt. }
		\label{fig:pollingVsum}
	\end{figure}
}

>>>>

\newcommand{\taskTwoHeatmap}{
	\begin{figure}
		\centering
		\includegraphics[width=0.9\columnwidth]{exp2histogram.pdf}
		\caption{2D histogram of placements from the prediction experiment. For each trial, participants were asked to place 5 tokens on heatmaps encoding both predicted safety, and uncertainty in these predictions. There were intentionally more tokens to place than there were ``safe and certain'' locations (top left corner). When looking at traditional 2D maps (left), participants favored safe but uncertain locations (bottom left corner). When looking at VSUPs (right), participants favored locations that were less safe, but more certain.)}
		\label{fig:taskTwoHeatmap}
	\end{figure}
}
%\newcommand{\performanceFig}{
%	\begin{figure}[t]
%		\centering
%		\includegraphics[width=0.9\columnwidth]{performance.pdf}
%		\caption{
%			The number of discrete color categories in traditional bivariate color maps versus VSUPs, assuming value is encoded by the Viridis color map~\protect\cite{viridis}, and uncertainty is encoded by saturation/value (whiter values are more uncertain). Since these whiter uncertain colors are closer together in CIELAB space, and VSUPs intentionally require fewer color bins in those regions, VSUPs always have more color categories than the corresponding 2D bivariate map. The red lines denote:
%		  \textbf{1)} Theoretical 50\% JND from CIELAB specification.
%			\textbf{2)} 50\% JND from controlled lab study of Mahy et al.~\protect\cite{mahy1994evaluation}.
%			\textbf{3)} 50\% JND from Szafir et al.\protect\cite{szafir2014adapting}.
%			\textbf{4)} 50\% JND for small visual targets from Stone et al.~\protect\cite{stone2014engineering}.
%			\textbf{5)} The threshold resulting in the simplest possible (2x2) bivariate map.
%	    }
%		\label{fig:performance}
%	\end{figure}
%}

@@

\newcommand{\pollFig}{
	\begin{figure}[t]
		\centering
		\includegraphics[width=\columnwidth]{polling-2d.pdf}
		\vspace{-15px}
		\caption{Polling data prior to the 2016 U.S. Presidential Election, encoded in a traditional 2D bivariate map. The redness and blueness show the polling lead for Trump and Clinton, respectively. But many polls had high margins of error, creating uncertainty about the election results.}
		\label{fig:polling2d}

@@

		\centering
		\includegraphics[width=\columnwidth]{polling-vsum.pdf}
		\vspace{-15px}
		\caption{Polling data prior to the 2016 U.S. Presidential Election, encoded with a VSUP. }
		\label{fig:pollingVsum}
	\end{figure}
}

### 2017-12-08 paper/main.tex
Added poll example

@@

We selected this task in order to promote risk-averse behavior. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In particular, there is greater perceived value in avoiding large losses as opposed to striving for a large gain~\cite{kahneman1979prospect}.

@@

 We recruited 24 participants for this task: $12$ female, $12$ male, ($M_{\text{age}}$= $37$, $SD_{\text{age}}$ = $9.8$). Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or wedge legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training and exclusion purposes. 3 people with unacceptably low accuracy on the training task ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$) were excluded from analysis.

@@

\subsubsection{Results} 
\taskTwoQuartiles

@@

The results partially support our first hypothesis. We performed a repeated measures ANOVA on our results to measure the effect of VSUP versus standard quantizations, and square- versus wedge-shaped legends, on average uncertainty in bets. We found no significant effect of quantization scheme on average guess uncertainty ($F(1,61)=0.05$, $p=0.83$). This result indicates that there does not appear to be a uniform pattern of risk aversion between the two scale types. However, given our quartile-based stimuli design, the central tendency of uncertainty would not necessarily capture caution in guesses. To capture differences in this non-normal distribution, and to test our belief that VSUP users would avoid highly uncertain regions, we performed a one-sided two-sample, Kolmogorov-Smirnov test on the distributions of uncertainty in guesses in the VSUP and traditional bivariate conditions. We found a significant difference ($D = 0.5$, $p=0.03$), and an inspection of the distribution shows that participants using VSUPs were less likely to make guesses with very high uncertainty. \figref{fig:taskTwoQuartiles} shows this result.

@@

@@

@@

VSUPs assume that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the virology dataset (\figref{fig:pollingVsum]}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.

>>>>

\pollFig

@@

We selected this task in order to promote risk-averse behavior. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In particular, there is greater perceived value in avoiding large losses as opposed to striving for a large gain~\cite{kahneman1979prospect}. Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. While other map types might produce different patterns of decision-making (for instance, the juxtaposed map might encourage participants to ignore uncertainty information altogether), their low accuracy for our previous information fusion tasks led us to discard them, as it would be difficult to distinguish between different patterns of predictions caused by the design, and different patterns caused by simply misreading the heatmap. Having fewer conditions also afforded a within-subjects design that controlled for the variation in inter-personal differences in strategies and risk-aversion, while limiting the potential effects of learning and fatigue on predictions from large numbers of stimuli.

@@

 We recruited 24 participants for this task: $12$ female, $12$ male, ($M_{\text{age}}$= $37$, $SD_{\text{age}}$ = $9.8$). Our selected square and wedge maps were either VSUPs or traditional bivariate maps, for a 2 (square or wedge legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training and exclusion purposes. 3 people with unacceptably low accuracy on the training task ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$) were excluded from analysis.

@@

\subsubsection{Results} 
\taskTwoHeatmap

@@

The results partially support our first hypothesis. We performed a repeated measures ANOVA on our results to measure the effect of VSUP versus standard quantizations, and square- versus wedge-shaped legends, on average uncertainty in bets. We found no significant effect of quantization scheme on average guess uncertainty ($F(1,61)=0.05$, $p=0.83$). This result indicates that there does not appear to be a uniform pattern of risk aversion between the two scale types. However, given our quartile-based stimuli design, the central tendency of uncertainty would not necessarily capture caution in guesses. To capture differences in this non-normal distribution, and to test our belief that VSUP users would avoid highly uncertain regions, we performed a one-sided two-sample, Kolmogorov-Smirnov test on the distributions of uncertainty in guesses in the VSUP and traditional bivariate conditions. We found a significant difference ($D = 0.5$, $p=0.03$), and an inspection of the distribution shows that participants using VSUPs were less likely to make guesses with very high uncertainty. \figref{fig:taskTwoHeatmap} shows this result.

@@

%exp1
%continuous bad
%juxtaposed bad

%exp2
%vsup: more cautious

@@

% results to discuss:
% bivariate maps: performance can be bad! esp. compared to univariate maps
% juxtaposition bad
% encourages caution
% different patterns of risk, same people

@@

VSUPs assume that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the polling dataset (\figref{fig:pollingVsum}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.
